---
title: "Computational Statistics"
subtitle: Lab 4
author: "Emil K Svensson and Rasmus Holm"
date: '`r Sys.Date()`'
output:
    pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
targetdensity <- function(x) {
    x^5 * exp(-x)
}

lognormalfuncs <- list(propsample=function(x) { rlnorm(1, meanlog=x, sdlog=1) },
                       propdensity=function(x, y) { dlnorm(x, meanlog=y, sdlog=1) },
                       targdensity=targetdensity)

chisquarefuncs <- list(propsample=function(x) { rchisq(1, df=floor(x + 1)) },
                       propdensity=function(x, y) { dchisq(x, df=floor(y + 1)) },
                       targdensity=targetdensity)

metropolis_hastings <- function(X0, iters, funcs) {
    x <- X0
    values <- rep(0, iters)

    alpha <- function(x, y) {
        numerator <- funcs$targdensity(y) * funcs$propdensity(x, y)
        denominator <- funcs$targdensity(x) * funcs$propdensity(y, x)
        numerator / denominator
    }

    for (i in 1:iters) {
        y <- funcs$propsample(x)
        u <- runif(1)

        if (u < alpha(x, y)) {
            x = y
        }

        values[i] <- x
    }

    values
}

iters <- 1000
X0 <- 1

actual <- rgamma(iters, shape=6, rate=1)
```

```{r}
set.seed(123456)
samples1 <- metropolis_hastings(X0=X0, iters=iters, funcs=lognormalfuncs)
mean(samples1)
plot(samples1, type="l")
```

```{r}
set.seed(123456)
samples2 <- metropolis_hastings(X0=X0, iters=iters, funcs=chisquarefuncs)
mean(samples2)
plot(samples2, type="l")
```

```{r}
x <- 0:20
y <- sapply(x, targetdensity)
plot(x, y)
```

```{r}
oldpar <- par(mfrow = c(1, 3))

hist(actual, main="Actual")
hist(samples1, main="Sampled (log-normal)")
hist(samples2, main="Sampled (chi-squared)")
```

```{r, echo=FALSE}
par(oldpar)
```


## Geldman rubin
```{r}
Geldman<-function(x){
  k <- nrow(x)
  n <- ncol(x)

  B <- (n / (k - 1)) * sum((rowMeans(x) - mean(x))^2)

  W <- sum((x - rowMeans(x))^2) / (k * (n - 1))

  VarV <- ((n - 1) / n) * W + B / n

  sqrtR <- sqrt(VarV / W)
  sqrtR
}

resultMatrix <- do.call(rbind, lapply(1:10, FUN = function(x)
    metropolis_hastings(X0 = x, iters = iters, funcs = chisquarefuncs)))

GeldmanRes <- Geldman(resultMatrix)
GeldmanRes
```


# Question 2

```{r}
load("../data/chemical.RData")

chem <- data.frame(X = X , Y = Y)


plot(chem)
hist(chem$Y)

hist(rbeta(50,3,2))
```

\begin{align*}
p(\mu) &= p(\mu_1) p(\mu_2 | \mu_1) \cdots p(\mu_n | \mu_{n-1}) \\
       &= \frac{1}{\sqrt{(2 \pi \sigma^2)^{n-1}}} \exp \left( - \frac{1}{2 \sigma^2} \sum_{i=2}^{n} (\mu_i - \mu_{i-1})^2  \right) \\
p(y | \mu) &= p(y_1 | \mu_1) p(y_2 | \mu_2) \cdots p(y_n | \mu_n) \\
           &= \frac{1}{\sqrt{(2 \pi \sigma^2)^{n}}} \exp \left( - \frac{1}{2 \sigma^2} \sum_{i=1}^{n} (y_i - \mu_i)^2 \right)
\end{align*}

\begin{align*}
p(\mu | y) &\propto p(y | \mu) p(\mu) \\
           &\propto \exp \left( - \frac{1}{2 \sigma^2} \left( \sum_{i=1}^{n} (y_i - \mu_i)^2 + \sum_{i=2}^{n} (\mu_i - \mu_{i-1})^2 \right) \right)
\end{align*}

This

\begin{align*}
p(\mu_1 | \mu_{-1}, y) &\propto \exp \left( -\frac{1}{\sigma^2} (\mu_1 - (y_1 + \mu_{2}) / 2)^2 \right) \\
p(\mu_i | \mu_{-i}, y) &\propto p(\mu_1 | \mu_{-1}, y) \exp \left( -\frac{3}{2 \sigma^2} \sum_{j=2}^{i} (\mu_j - (y_j + \mu_{j-1} + \mu_{j+1}) / 3)^2 \right) \\
p(\mu_n | \mu_{-n}, y) &\propto p(\mu_{n-1} | \mu_{-n-1}, y) \exp \left( -\frac{1}{\sigma^2} (\mu_n - (y_n + \mu_{n-1}) / 2)^2 \right) \\
\end{align*}

Alternatively this

\begin{align*}
p(\mu_1 | \mu_{-1}, y) &\propto \exp \left( -\frac{1}{\sigma^2} (\mu_1 - (y_1 + \mu_{2}) / 2)^2 \right) \\
p(\mu_i | \mu_{-i}, y) &\propto \exp \left( -\frac{3}{2 \sigma^2} (\mu_i - (y_i + \mu_{i-1} + \mu_{i+1}) / 3)^2 \right) \\
p(\mu_n | \mu_{-n}, y) &\propto \exp \left( -\frac{1}{\sigma^2} (\mu_n - (y_n + \mu_{n-1}) / 2)^2 \right) \\
\end{align*}
